Binning Strategies and Related Loss for Large Data

### 2013-03-30 Paper.Rnw
major overhaul of the baseball loss plots, color scheme still needs some work, surrounding paragraphs to be edited this evening

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)

@@

<<recover,echo=FALSE,results='asis'>>=
set.seed(90)
x <- rnorm(100, 5, 2)
y <- rnorm(100, 5, 2)
d <- data.frame(x=x, y=y, c=1)
d <- d[order(d$x, d$y),]
print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
dbin <- dbin[order(dbin$x, dbin$y),]
print(xtable(head(dbin), digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')
names(d) <- c("x", "y", "Freq")
drecover <- unique(lossCalc(d, c(1,1), "standard", newData=TRUE)$NewData)
names(drecover) <- c("x", "y", "c")
drecover <- drecover[order(drecover$x, drecover$y),]
print(xtable(head(drecover), digits=c(0, 4, 4, 0), caption=paste('Recovered Data\\newline', nrow(drecover), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
@
\begin{table}[hbtp]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 100 rows]{
\begin{tabular}{rrr}\hline
       $x$ &      $y$ & c \\ \hline
 -0.2582 & 1.5497 & 1 \\ 
 -0.3843 & 5.6508 & 1 \\ 
 1.0597 & 2.4308 & 1 \\ 
 2.2680 & 1.5877 & 1 \\ 
 2.4992 & 4.6260 & 1 \\ 
 1.6788 & 5.0260 & 1 \\ 
 1.6054 & 6.4125 & 1 \\ 
 2.4180 & 6.4438 & 1 \\ 
 2.4345 & 6.4462 & 1 \\ 
 2.3620 & 10.3119 & 1 \\  \hline
\end{tabular}}\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Visual Centers, 49 rows]{
\begin{tabularx}{.7\textwidth}{XXc} \hline
  $x^\ast$ & $y^\ast$ & c \\ \hline
  0 & 2 & 1 \\ 
  0 & 6 & 1 \\ 
  1 & 2 & 1 \\
  2 & 2 & 1 \\
  2 & 5 & 2 \\ 
  2 & 6 & 3 \\ 
  2 & 10 & 1 \\ 
   \hline
\end{tabularx}
}\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Numerical Centers, 49 rows]{
\begin{tabular}{rrc}\hline
  $\bar{x}$   &    $\bar{y}$ & c \\ \hline
 -0.3843 & 5.6508 & 1 \\ 
 -0.2582 & 1.5497 & 1 \\ 
  1.0597 & 2.4308 & 1 \\ 
  2.2680 & 1.5877 & 1 \\  
  2.0890 & 4.8260 & 2 \\ 
  2.1526 & 6.4341 & 3 \\ 
  2.2680 & 1.5877 & 1 \\ \hline
\end{tabular}
}\end{minipage}
\caption{\label{RecoveryComparison}ten rows from Original and Binned Data Tables, with data storage sizes}
\end{table}

@@

 These characteristics of the numerical and visual losses are shown using the baseball data in the total combined visual and numerical loss at different bin widths in $x$ (Games) in Figure ~\ref{fig:LossComparison1}.

\begin{figure}[hbtp]
\centering
<<LossComparison1, echo=FALSE, eval=T, fig.width=6, fig.height=4, out.width='.75\\linewidth', out.height='.495\\linewidth'>>=

losses <- read.csv("../dbData/data/losses.csv", stringsAsFactors=FALSE)

qplot(x=x, y=TotalLoss.G*100, group=interaction(y, Algorithm), data=subset(losses, y<=10), geom="line", colour=Algorithm, ylab="Percent Spatial Loss", xlab="Bin  Width  (Games)")+ theme_bw() + theme(legend.position="bottom") + ggtitle("Total Combined Numerical and Visual Loss:  Baseball Games") + guides(colour = guide_legend(override.aes = list(size = 2))) + facet_wrap(~Algorithm)
@
\caption{\label{fig:LossComparison1} Spatial Loss increases with bin width for each variable. The random algorithm has higher spatial loss than the standard algorithm.}
\end{figure}
The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary. The standard algorithm is preferable because it has lower total combined numerical and visual loss, but the two algorithms are very similar when we compare frequency loss. The random algorithm does have more variability, but the frequency loss values are much more similar, and when bins are of size two in either or both dimensions, the random algorithm has lower frequency loss than the standard algorithm. This holds for bins which are powers of two, as long as the random binning algorithm is applied iteratively to take advantage of the constant loss property noted above. %~\ref{fig:LossComparison}

<<LossComparison, echo=FALSE, fig.cap="Frequency Loss increases as bin width increases in Games and Strike Outs. The random binning algorithm has slightly higher frequency loss as bin width grows large, but the loss is comparable when bin widths are small.", fig.width=8, fig.height=4, out.width='\\linewidth', out.height='.5\\linewidth', fig.pos='H'>>=
# 
# connect <- dbConnect(dbDriver("MySQL"), user="2009Expo", 
#                      password="R R0cks", port=3306, dbname="baseball", 
#                      host="headnode.stat.iastate.edu")
# pitch <- new("dataDB", co=connect, table="Pitching")
# d1 <- dbData(pitch, vars=c( "G", "SO"))
# bins <- expand.grid(x=c(1:10, 15, 20), y=c(1:5, seq(10, 100, 15)))
# 
# library(multicore)
# lossesRdm <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="random")))
# lossesStd <- do.call("rbind", mclapply(1:nrow(bins), function(i) lossCalc(data=d1, binning=c(bins[i,1], bins[i,2]), type="standard")))
# 
# lossesRdm <- cbind(bins, lossesRdm)
# lossesStd <- cbind(bins, lossesStd)
# 
# losses <- rbind(cbind(Algorithm="Random", lossesRdm), cbind(Algorithm="Standard", lossesStd))
# write.csv(losses, "../dbData/data/losses.csv", row.names=FALSE)
losses <- read.csv("../dbData/data/losses.csv", stringsAsFactors=FALSE)

qplot(x=y, y=TotalLoss.LogFreq, group=interaction(x, Algorithm), data=losses, geom="line", colour=Algorithm, ylab="Frequency Loss", xlab="Bin Width (Strike Outs)")+ facet_grid(.~Algorithm) + scale_colour_discrete(guide="none") + theme_bw() + ggtitle("Frequency Loss: Baseball Games and Strike-Outs")
@

@@

For this data example we will revisit the baseball data used earlier in this paper.

>>>>

\section{Introduction}
<<setup,echo=F,include=F,eval=T>>=
library(ggplot2)
library(gridExtra)
library(dbData)
library(reshape2)
#library(multicore)
library(devtools)
library(xtable)
library(RColorBrewer)

@@

<<recover,echo=FALSE,eval=FALSE,results='asis'>>=
set.seed(90)
x <- rnorm(100, 5, 2)
y <- rnorm(100, 5, 2)
d <- data.frame(x=x, y=y, c=1)
d <- d[order(d$x, d$y),]
print(xtable(head(d,10)[c(2,1,3,6,10,5,4,8,9,7),], digits=c(0, 4, 4, 0), caption=paste('Original Data\\newline', nrow(d), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
dbin <- ddply(binStd(d, c(1,1)), .(x,y), summarize, c=sum(c))
dbin <- dbin[order(dbin$x, dbin$y),]
print(xtable(head(dbin), digits=c(0, 0, 0, 0), caption=paste('Binned Data\\newline', nrow(dbin), 'rows'), align=c('X','X','X','c')), include.rownames=FALSE, table.placement="hbtp", floating=FALSE, tabular.environment="tabularx", width='.7\\textwidth')
names(d) <- c("x", "y", "Freq")
drecover <- unique(lossCalc(d, c(1,1), "standard", newData=TRUE)$NewData)
names(drecover) <- c("x", "y", "c")
drecover <- drecover[order(drecover$x, drecover$y),]
print(xtable(head(drecover), digits=c(0, 4, 4, 0), caption=paste('Recovered Data\\newline', nrow(drecover), 'rows'), include.rownames=FALSE, table.placement="hbtp", floating=FALSE))
@

\begin{table}[hbtp]
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Original Data, 100 rows]{
\begin{tabular}{rrr}\hline
       $x$ &      $y$ & c \\ \hline
 -0.2582 & 1.5497 & 1 \\ 
 -0.3843 & 5.6508 & 1 \\ 
 1.0597 & 2.4308 & 1 \\ 
 2.2680 & 1.5877 & 1 \\ 
 2.4992 & 4.6260 & 1 \\ 
 1.6788 & 5.0260 & 1 \\ 
 1.6054 & 6.4125 & 1 \\ 
 2.4180 & 6.4438 & 1 \\ 
 2.4345 & 6.4462 & 1 \\ 
 2.3620 & 10.3119 & 1 \\  \hline
\end{tabular}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Visual Centers, 49 rows]{
\begin{tabularx}{.7\textwidth}{XXc} \hline
  $x^\ast$ & $y^\ast$ & c \\ \hline
  0 & 2 & 1 \\ 
  0 & 6 & 1 \\ 
  1 & 2 & 1 \\
  2 & 2 & 1 \\
  2 & 5 & 2 \\ 
  2 & 6 & 3 \\ 
  2 & 10 & 1 \\ 
   \hline
\end{tabularx}}
\end{minipage}\hfil
\begin{minipage}[t]{.33\textwidth}\centering\subfloat[Binned Data Numerical Centers, 49 rows]{
\begin{tabular}{rrc}\hline
  $\bar{x}$   &    $\bar{y}$ & c \\ \hline
 -0.3843 & 5.6508 & 1 \\ 
 -0.2582 & 1.5497 & 1 \\ 
  1.0597 & 2.4308 & 1 \\ 
  2.2680 & 1.5877 & 1 \\  
  2.0890 & 4.8260 & 2 \\ 
  2.1526 & 6.4341 & 3 \\ 
  2.2680 & 1.5877 & 1 \\ \hline
\end{tabular}}
\end{minipage}
\caption{\label{RecoveryComparison}ten rows from Original and Binned Data Tables, with data storage sizes}
\end{table}

@@

The far more important factor is the frequency loss, as we perceive that loss in the variability of the shading of bins which has a larger effect on the utility of the graphical data summary. The standard algorithm is preferable because it has lower total combined numerical and visual loss, but the two algorithms are very similar when we compare frequency loss. The random algorithm does have more variability, but the frequency loss values are much more similar, and when bins are of size two in either or both dimensions, the random algorithm has lower frequency loss than the standard algorithm. This holds for bins which are powers of two, as long as the random binning algorithm is applied iteratively to take advantage of the constant loss property noted above. %~\ref{fig:LossComparison}

@@

For this data example we will revisit the baseball data used earlier in this paper.  The characteristics of loss described in the discussion above are exemplified by viewing the loss for the baseball data in Figure ~\ref{fig:LossComparison1}.

\begin{figure}[hbtp]
\centering
<<LossComparison1, echo=FALSE, eval=T, fig.width=6.5, fig.height=8>>=

losses <- read.csv("./data/losses.csv", stringsAsFactors=FALSE)
lossmelt <- melt(losses[,c(1:4,6,10)],id=c("Algorithm","x","y"))
lossmelt <- lossmelt[which(lossmelt$y <= 10),]
lossmelt$y <- as.factor(lossmelt$y)
levels(lossmelt$variable) <- c("Numerical Loss","Visual Loss", "Log Frequency Loss")
qplot(x=x, y=value,
      group=interaction(y, Algorithm),
      data=lossmelt, geom="line",
      colour=y,
      xlab="Bin  Width  (Games)", ylab="Proportion Loss")+   
        theme_bw() + theme(legend.position="bottom") + 
        guides(colour = guide_legend(title="Bin Width (Strike Outs)", override.aes = list(size = 2))) +
        facet_grid(variable ~ Algorithm, scales = "free")+
        ggtitle("Loss for Baseball Data under Standard and Random Binning")

@
\caption{\label{fig:LossComparison1} Loss funcion }
\end{figure}
